Spit-Apply-Combine with Collapsing Groups

### 2022-11-18 paper/paper.tex
some work
updates and improvements

@@

@@

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(a))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
Here, the output is collected in a set $R$. We also introduced some new
notation: the notation $2^U$ stands for `the set of all subsets of $U$'. The
notation $f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you
give it a set of labels from $A$ and it returns all the elements of $U$ that
have any of those labels. This definition is slightly too general for the
current algorithm, but it will come in handy when we move to collapsing groups.

@@

\subsection{Split-apply-combine with collapsing groups}
The goal of the algorithm is to compute a value for each 
subset of a dataset, possibly using values from a larger
subset. The input of the algorithm consists again of a
finite set $U$, and aggregation function $\phi$ that takes
a subset of $U$ and returns a value in some domain $X$.
Compared to Algorithm~\ref{alg:sac} we need two extra
inputs. First, we need a user-defined function that checks whether
a given subset $d$ of $U$ is suitable for computing $\phi(d)$.
We will call this function $\beta: 2^U\to \mathbb{B}$, where
$\mathbb{B}=\{\texttt{True},\texttt{False}\}$. 
Typical tests are checking whether there are sufficient records
available, or whether certain variables have a low enough fraction
of missing values. Second, we need
a \emph{collapsing scheme}
\begin{equation}
U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

@@

To generalize this algorithm to generic collapsing schemes, we introduce
the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
returns the corresponding label in $A_k$. This can formally be defined 
as consecutive application of $f_1, f_2\cdots, f_k$ to an element of 
$A_0$. In notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to that label via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally
write down our main Algorithm~\ref{alg:saccg}.

@@

\Output{$R$: the value of $\phi$ for every part of $U$, for which a suitable
collapsing group can be found, as a set of triples $(a,k,x)\in A_0\times
A_k\times X $ with $0\leq k\leq n$.}

$R = \{\}$\;
\For{$a\in A_0$}{
  $i=0$\;
  $d = f_0^{-1}(\{a\})$\tcp*{get subset of $U$}
  \While{$i<n \land \lnot\beta(d)$}{
    $i = i+1$ \tcp*{Increase collapse level} 
    $d = (f_0^{-1}\circ F_i^{-1}\circ F_i)(a)$ \tcp*{Collapse and get subset}
  }
  \If{$\beta(d)$}{ \label{line:cond}
    $R = R\cup \{(a, i,\phi(d))\}$\;
  } 
}

@@

\section{The \texttt{rcollapse} package}

\begin{itemize}
\item{Specification of collapsing sequence with multiple collapsing variables (e.g. first size class, then NACE code)}
\end{itemize}

>>>>

\usepackage{colortbl, xcolor}

@@

\begin{itemize}
\item Add SDC collapsing as application
\end{itemize}

@@

$R = \{\}$\;
\For{$a\in A$}{
  $d=f^{-1}(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
Here, the output is collected in a set $R$. We also introduced some new
notation: the notation $2^U$ stands for `the set of all subsets of $U$'. The
notation $f^{-1}$ stands for the formal \emph{inverse} of $f$. That is: you
give it a set of labels from $A$ and it returns all the elements of $U$ that
have any of those labels. This definition is slightly too general for the
current algorithm, but it will come in handy when we move to collapsing groups.

@@

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each 
subset of a dataset, possibly using values from a larger
subset. The input of the algorithm consists again of a
finite set $U$, and aggregation function $\phi$ that takes
a subset of $U$ and returns a value in some domain $X$.
Compared to Algorithm~\ref{alg:sac} we need two extra
inputs. First, we need a user-defined function that checks whether
a given subset $d$ of $U$ is suitable for computing $\phi(d)$.
We will call this function $\beta: 2^U\to \mathbb{B}$, where
$\mathbb{B}=\{\texttt{True},\texttt{False}\}$. 
Typical tests are checking whether there are sufficient records
available, or whether certain variables have a low enough fraction
of missing values. Second, we need
a \emph{collapsing scheme}
\begin{equation}
U\xrightarrow{f_0}A_0\xrightarrow{f_1}A_1\xrightarrow{f_2}\cdots\xrightarrow{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

@@

To generalize this algorithm to generic collapsing schemes, we introduce
the function $F_k:A_0\to A_k$, which accepts a label in $A_0$ and 
returns the corresponding label in $A_k$. This can formally be defined 
as consecutive application of $f_1, f_2,\cdots, f_k$ to an element of 
$A_0$. In notation
\begin{equation*}
F_k \equiv f_k\circ f_{k-1}\circ\cdots\circ f_1.
\end{equation*}
Similarly we define $F^{-1}_k:2^{A_k}\to 2^{A_0}$ as the composition
\begin{equation*}
F_k^{-1} \equiv f_1^{-1}\circ f_2^{-1}\circ\cdots\circ f_k^{-1}.
\end{equation*}
This function asks for a set of labels in $A_k$ and returns all the labels in
$A_0$ that are mapped to that label via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  We are now ready to formally
write down our main Algorithm~\ref{alg:saccg}.

@@

\Output{$R$: the value of $\phi$ for every part of $U$, for which a suitable
collapsing group can be found, as a set of triples $(a,k,x)\in A_0\times
\underline{n}\times X $ where $\underline{n}=\{0,1,\ldots,n\}$.}

$R = \{\}$\;
\For{$a\in A_0$}{
  $i=0$\;
  $d = f_0^{-1}(\{a\})$\tcp*{get subset of $U$}
  \While{$i<n \land \lnot\beta(d)$}{
    $i = i+1$ \tcp*{Increase collapse level} 
    $d = (f_0^{-1}\circ F_i^{-1}\circ F_i)(a)$ \tcp*{Collapse and get subset}
  }
  \If{$i<n \lor \beta(d)$}{ \label{line:cond}
    $R = R\cup \{(a, i,\phi(d))\}$\;
  } 
}

@@

\section{The \texttt{rcollapse} package}
Grouped aggregation with a fall-though scenario based on a collapsing scheme
requires a lot of parameters to be specified by the user. Besides the data, one
needs to specify the method(s) of aggregation, the collapsing scheme, and the
condition to decide whether a subset is fit for aggregation or a next collapse
is necessary. The main function of the package is caled \texttt{collapse}
and has the following signature.
\begin{verbatim}
  collapse(data, collapse, test, ...)
\end{verbatim}
where \texttt{data} is a data frame, \texttt{collapse} represents the collapse
sequence, \texttt{test} is a function that accepts a subset of \texttt{data}
and returns \texttt{TRUE} or \texttt{FALSE} and the ellepsis is a sequence
of \texttt{label = expression} pairs that represent aggregation
functions, in a similar style as \texttt{dplyr::summarize}. The output
is a data frame that can schematically be represented as follows.
\begin{verbatim}
  [Grouping Variables, Collapse level, Output aggregates]
\end{verbatim}
The first columns represent the variables that define the output grouping, the
next column is an integer that indicates the level of collapsing used to
compute the aggregate, and the last set of columns store the aggregates.

To accommodate the different use cases, the \texttt{rcollapse} package supports
two different interfaces for specifying collapse sequences.  The first and most
general is the formula interface, that requires that the collapsing sequence is
represented as variables in the data set to be aggregated. The second is the
tabular interface discussed before. This only supports a single grouping
variable. For hierarchical classifications there is a helper function to
translate parent-child representations of such classifications into a
collapsing sequence.

Besides complicated collapsing schemes, users must have the ability to specify
a possibly complex condition for determining when a dataset is suitable for
aggregation or not. The package offers some convenience functions that support common
use cases as well as support for specifying generic conditions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The formula interface}
It is not uncommon for users to use multiple variables for defining collapsing
schemes. For example when grouping by two categorical variables \texttt{A} and
\texttt{B}, one may first want to collapse variable $\texttt{B}$ into a
variable \texttt{B1} and if that is not sufficient, in the next step grouping
in the second variable is dropped so grouping is only defined by \texttt{A}.
In \texttt{rcollapse} such a scheme is denoted with the formula \texttt{A*B \~{}
A*B1 + A}. Here, the left-hand-side denotes the desired output grouping level
that is attempted first, and the right-hand-side denotes consecutive collapse
levels separated by \texttt{+}.

The following tables illustrate the idea. The left table represents the data
set to be aggregated by \texttt{A} and \texttt{B}. In this example we wish to
compute the sum $t_Y$ of column $Y$ per group. We demand that there are at
least three records in a group, otherwise we collapse one step.
%
\begin{center}
\begin{tabular}{cccl}
\multicolumn{4}{l}{\texttt{Input data}}\\
\hline
\texttt{A} & \texttt{B} & \texttt{B1} & $Y$\\
\hline
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   1   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   2   \\
\cellcolor{red!25}   1 &\cellcolor{red!25}   11 &                1       & \cellcolor{red!25}   4   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 8   \\
\cellcolor{green!25} 2 &                     12 & \cellcolor{green!25} 1 & \cellcolor{green!25} 16  \\
\cellcolor{green!25} 2 &                     13 & \cellcolor{green!25} 1 & \cellcolor{green!25} 32  \\
\cellcolor{blue!25}  3 &                     21 &                2       & \cellcolor{blue!25}  64  \\
\cellcolor{blue!25}  3 &                     22 &                2       & \cellcolor{blue!25}  128 \\
\cellcolor{blue!25}  3 &                     12 &                1       & \cellcolor{blue!25}  256 \\
\hline
\end{tabular}\hspace{1cm}\begin{tabular}{ccl}
\multicolumn{3}{l}{\texttt{Output}}\\
\hline
\texttt{A*B} & \texttt{level} & $t_Y$\\
\hline
\rowcolor{red!25}   1 11        & 0             & 7  \\
\rowcolor{green!25} 2 12        & 1             & 56 \\
\rowcolor{green!25} 2 13        & 1             & 56 \\
\rowcolor{blue!25}  3 21        & 2             & 448\\
\rowcolor{blue!25}  3 22        & 2             & 448\\
\rowcolor{blue!25}  3 12        & 2             & 448\\
\hline
\end{tabular}
\end{center}
%
The table on the right represents the output. Colors indicate which data was
used. The First row in the output represents group
$(\texttt{A}=1,\texttt{B}=11)$. The collapsing level is zero, which means
that no collapsing was necessary. Indeed, in the data table we see that there
are three rows with \texttt{A==1} and \texttt{B==11} with $Y$ values $1, 2$,
and $7$, resulting in $t_Y=7$ for this group.

Next, the algorithm tries to compute the total for group
$(\texttt{A}=2,\texttt{B}=12)$ (in green) but finds that there are only two
such rows. It now defines a new group $(\texttt{A}=2,\texttt{B1}=1)$ and finds
that there are three records in that group so it computes $t_Y=8+16+32=56$.
Similarly, it finds one record with $(\texttt{A}=2,\texttt{B}=13)$ and
collapses to $(\texttt{A}=2,\texttt{B1}=1)$, again yielding $t_y=56$.

Finally, for $(\texttt{A}=2,\texttt{B}=21)$ the algorithm finds only a single
record. It collapses to $(\texttt{A}=2,\texttt{B1}=2)$ and finds two so it
collapses further to $(\texttt{A}=2)$ to find three records.  This yields
$t_Y=64+128+256=448$. Similarly, the groups $(\texttt{A}=3, \texttt{B}=22)$ and
$(\texttt{A}=3,\texttt{B}=12)$ are collapsed to $(\texttt{A}=2)$.

In R code the previous example can be executed as follows.
\begin{verbatim}
  > library(rcollapse)
  > input <- data.frame(
  +      A  = c(1,1,1,2,2,2,3,3,3)
  +    , B  = c(11,11,11,12,12,13,21,22,12)
  +    , B1 = c(1,1,1,1,1,1,2,2,1)
  +    , Y  = 2^(0:8)
  + )
  >
  > collapse(data     = input
  +        , collapse = A*B ~ A*B1 + A
  +        , test     = function(d) nrow(d)>=3
  +        , tY       = sum(Y) )
    A  B level  tY
  1 1 11     0   7
  2 2 12     1  56
  3 2 13     1  56
  4 3 21     2 448
  5 3 22     2 448
  6 3 12     2 448
\end{verbatim}
%
The formula interface allows users to quickly experiment with different
collapsing schemes, but only after the necessary variables have been added to
the original data frame, which may require some extensive data manipulation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The data frame interface}

\subsection{Specifying tests}

\subsection{Programming over \texttt{collapse}}

\section{Extensive example: imputing economic data}

\section{Summary and conclusion}

\begin{itemize}
\item{Record-wise results}
\item{Record-wise imputation}
\item{Representation of hierarchical classifications: parent-child lists vs tabular}
\end{itemize}
